---
title: "CSCI 456 Group 2 - Assignment #2"
author: "Jacob Collins, Naina K."
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document: default
  html_document: default
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse) # select
library(dplyr) # Mutate
library(ggplot2) # Plot
default_theme <- theme(
    # Title
    plot.title = element_text( # Title
      color = "white", size = 14,  
      face = "bold", hjust = 0.5),
    plot.subtitle = element_text( # Subtitle
      color = "lightgray", size=10, 
      face = "italic", hjust = 0.5),
    # Axis Text
    axis.title.x = element_text( # X-Axis Title
      color = "lightgray",
      size = 12, angle = 0,),
    axis.title.y = element_text( # Y-Axis Title
      color = "lightgray",  
      size = 12, angle = 90,),
    axis.text.x = element_text( # X-Axis Labels
      face = "italic", color = "lightgray",  
      size = 8, angle = 0),
    axis.text.y = element_text( # Y-Axis Labels
      face = "italic", color = "lightgray",  
      size = 8, angle=90),
#    axis.text.x = element_text( # X-Axis Labels
#      face = "italic", color = "lightgray",  
#      size = 8, angle = -15),
#    axis.text.y = element_text( # Y-Axis Labels
#      face = "italic", color = "lightgray",  
#      size = 8, angle=45),
    # Background Colors
    panel.background = element_rect( # Background of the Plot Box Itself
      fill = "#060606", colour = "#5f5d63",
      linewidth = 2, linetype = "solid"),
    panel.grid.major = element_line( # Grid Colors
      linewidth = 0.5, linetype = 'solid', 
      colour = "#7a797a"), 
#    panel.grid.minor = element_line( # Minor Grid Colors
#      linewidth = 0.25, linetype = 'solid', 
#      colour = "#585759"),
    plot.background = element_rect(fill = "#0e0e0f"), # Furthest Background
    # Legend
    legend.title = element_text(color = "lightgray", size = 10),
    legend.text = element_text(color = "gray"),
    legend.key = element_rect(fill = "#383838", color = NA),
    #legend.key.size = unit(1.5, "cm"),
    #legend.key.width = unit(0.5,"cm"), 
    legend.background = element_rect(fill = "#5f5d63"),
  )
```

# Data Exploration and Cleaning

## Select a Data Source:

[Department of Agriculture- National USFS Fire Occurrence Point](https://catalog.data.gov/dataset/national-usfs-fire-occurrence-point-feature-layer-d3233)

## Formulate Research Questions:

Develop 3-5 research questions that you aim to answer through your analysis. Ensure that these questions are specific, measurable, and relevant to the dataset you have chosen.

- How has the size and frequency of fires changed over time?
- What regions are more susceptible to forest fires, in terms of longitude and latitude?
- What is the relationship between wildfire occurrence and proximity to human settlements? 


## Data Variable Description:

Explain the variables in your dataset you intend to analyze.

- `STATCAUSE` - Cause: indicates cause of fire , categorical-qualitative data
- `TOTALACRES` - Total acres: represents area burned by the wildfire, quantitative- continuous data
- `LATDD83` - Latitude: north-south location, quantitative- continuous data
- `LONGDD83` - Longitude: east-west location, quantitative- continuous data
- `REVDATE` - Fire year: year wildfire occurred, quantitative- discrete data


## Preparation:

Describe how you did data preparation. This may include: - Checking and handling missing values - Encoding categorical variables

### Initial csv read

```{r cars}
fires <- read.csv("../data/fires.csv", fill=TRUE)
fires <- fires %>% select(TOTALACRES, LATDD83, LONGDD83, STATCAUSE, REVDATE)
nrow(fires)
summary(fires)
```

There are 582,034 observations.

### Cleaning `REVDATE`

```{r}
fires$REVDATE <- as.Date(fires$REVDATE)
summary(fires$REVDATE)
```

It seems there are many years incorrectly labeled.

```{r}
# No real data in the future
fires$REVDATE[as.numeric(format(fires$REVDATE, "%Y")) > 2025] <- NA
# No real data before 1500 is a safe assumption
fires$REVDATE[as.numeric(format(fires$REVDATE, "%Y")) < 1500] <- NA

fires$REV.YEAR <- format(fires$REVDATE, "%Y")
fires$REV.YEAR <- as.numeric(fires$REV.YEAR)
fires %>% ggplot(aes(x = REV.YEAR)) +
  geom_histogram(binwidth=1, fill = "steelblue", color="white", na.rm=TRUE) +
  default_theme +
  labs(
    x = "Year",
    y = "Count",
    title = "Count of Fires by Year",
  ) + theme(
    axis.text.x = element_text( # X-Axis Labels
      face = "italic", color = "lightgray",  
      size = 8, angle = 90)
  )
```

Viewing this plot shows us where the bulk of the data is, and we can prune around it.

```{r}
# This deleted REVDATE and filled around Y, M, & D with the current md, yd, ym, respectively.
#fires <- fires %>% separate(REVDATE, into = c("REV.Y", "REV.M", "REV.D"), sep = "-")
#fires$REV.Y <- as.Date(fires$REV.Y, format = "%Y")
#fires$REV.M <- as.Date(fires$REV.M, format = "%m")
#fires$REV.D <- as.Date(fires$REV.D, format = "%d")
#summary(fires %>% select(REV.Y, REV.M, REV.D))

# No real data before 1985
fires$REVDATE[as.numeric(format(fires$REVDATE, "%Y")) < 1985] <- NA
summary(fires$REVDATE)
```

```{r}
fires$REV.YEAR <- format(fires$REVDATE, "%Y")
fires$REV.YEAR <- as.numeric(fires$REV.YEAR)
fires %>% ggplot(aes(x = REV.YEAR)) +
  geom_histogram(binwidth=1, fill = "steelblue", color="white", na.rm=TRUE) +
  default_theme +
  labs(
    x = "Year",
    y = "Count",
    title = "Count of Fires by Year",
  ) + theme(
    axis.text.x = element_text( # X-Axis Labels
      face = "italic", color = "lightgray",  
      size = 8, angle = 90)
  )
```

### Cleaning Acres

```{r}
summary(fires$TOTALACRES)
```

```{r}
fires %>% ggplot(mapping=aes(x=REVDATE, y=TOTALACRES)) +
  geom_point(color="white", alpha=0.3, na.rm=TRUE) +
  default_theme +
  labs(
    x = "Date",
    y = "Total Acres",
    title = "Size of Fires by Date"
  )
```

Acres looks alright.

### Cleaning Longitude and Latitude

```{r}
summary(fires %>% select(LATDD83, LONGDD83))
```

The valid range for latitude is -90 to 90. The valid range for longitude is -180 to 180.

```{r}
fires$LATDD83[fires$LATDD83 > 90] <- NA
fires$LATDD83[fires$LATDD83 < -90] <- NA
fires$LONGDD83[fires$LONGDD83 > 180] <- NA
fires$LONGDD83[fires$LONGDD83 < -180] <- NA
summary(fires %>% select(LATDD83, LONGDD83))
```


### Cleaning `STATCAUSE`

```{r}
unique(fires$STATCAUSE)
```

These can be joined by matching number or obvious typos like two spaces instead of one. 

Blank results, miscellaneous, undet., etc. shall be marked NA.






















